Shader Pipeline and Effect Encapsulation using CLOS

### 2019-01-30 els2019-shader-pipeline/paper.tex
Fix the figures slightly, !@#/# TEX!!!
Incorporate Robert Strandh's feedback

\begin{abstract}
  Modern real-time graphics make use of a lot of tricks in order to produce stunning visuals. Many of these tricks require separate rendering passes, as well as separate rendering logic for each pass. These passes are then combined in a variety of ways in order to produce a final image. The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex. Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with either the object or the pass. We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance.
\end{abstract}

@@

\def\abovecaptionskip{1pt}
\def\listingautorefname{listing}
\def\figureautorefname{figure}

@@

Each of the customisable stages only accept a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour. Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute. \\

@@

This results in a difficult challenge for modularity. We attempt to solve this challenge through several systems:

@@

The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised in a very limited extent with custom shaders. These shaders must fit into Unity's existing lighting and overall rendering model. While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features. \\

The work by He et al.\cite{components} introduces a framework for general encapsulation of shaders and their parameters into structures that minimise the overhead of changing GPU state while retaining the ability to dynamically compose shader parts. They do however not create a distinction between properties for the rendering of an object, and those for the rendering of an overall pass. \\

@@

In our previous work\cite{glsloop} we introduce a system to tie shader code to classes and compose behaviour through inheritance. We make use of this system and extend it to allow further control over rendering behaviour in individual passes.

\section{Overview}\label{overview}
The system is composed out of three distinct entities: shader objects, shader passes, and pipelines. Both shader objects and shader passes encapsulate rendering logic. The pipeline on the other hand represents the assembly of a complete frame and only influences rendering in the sense that it tracks which passes to run in what order. \\

Due to the restrictions of OpenGL's rendering pipeline, the combination of the render logic of objects and passes is not trivial. We present a protocol to solve this in \hyperref[passes]{section 4}. \\

In \hyperref[pipelines]{section 5} we present a protocol to represent input and output information of a pass and how to connect these inputs and outputs between passes together. \\

@@

\section{Passes}\label{passes}
A shader pass should encapsulate the logic for drawing objects in a certain way. In order to do this, the pass needs to be able to partially or fully control the shaders of objects. For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it. The shader pass should be able to make use of this information if it needs to, or discard it completely if unneeded. \\

In order to allow this we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}. This technique allows us to combine pieces of GLSL code. On top of this shader combination sits a new protocol to control the interaction between shader passes and objects. \\

@@

For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called. This allows the pass to prepare the shader program that will be used during rendering for this object. A shader program is an OpenGL resource that compiles the shaders for the stages of a pass together. Using \\ \code{shader-program-for-pass} this program can then later be retrieved. The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code. \\

\code{register-object-for-pass} calls \code{make-pass-shader-program} to compute this program and registers it internally in the shader pass so that it can be retrieved later. If the shader pass would like to retain complete control over how each object is drawn, it would forego this call and instead generate its own shader program. This is especially useful for post-processing effects that don't render any objects at all, and instead simply operate on textures output by previous passes. \\

@@

\code{coerce-pass-shader} computes the effective shader source for a particular shader stage or type. Typically this simply involves the combination of the shader sources of both shader pass and object for the given shader type. We perform this combination using the same parsing and code walking strategy as we described in our previous work.\cite{glsloop} \\

@@

This function is fine for allowing the object control over the behaviour. However, the pass cannot exert the same amount of control due to the generic function's argument precedence. For example, if an \code{:around} method specialised on the object exists, a pass would not have any way of preventing it from firing, as its own \code{:around} method would be executed afterwards. We thus introduce another function with inverted argument order that is called first. \\

@@

An example of such a pass is illustrated in \autoref{lst:deferred-render}. The ports are modelled as slots on the class, requiring a new metaclass. A slot with a port-type will not only hold a texture for its value, but also retain information about how it is connected to other passes. Using slots in this manner also allows us to inherit ports from other classes, and thus combine behaviour alongside the shader source code. \\

@@

When assembling a pipeline, we can then connect these ports together in order to dictate how the various output textures generated by a pass are used as inputs for other passes. As an example, the pipeline from \autoref{fig:render-pipeline} is created in \autoref{lst:pipeline}. \\

@@

\begin{step}
\item Create a pipeline object and instances of the desired shader passes.
\item Connect the ports of the shader passes.
\item Prepare the pipeline to allocate the needed resources.
\item Create instances of the desired objects.
\item Register each object with each pass.
\item Call \code{render} on the pipeline with a collection of all objects to draw.
\item The output port's texture of the last shader pass in the pipeline will contain the finished frame.
\end{step}

@@

\section{Allocation}\label{allocation}
In order to prepare the shader pipeline, several resources need to be allocated. Each shader stage needs a ``framebuffer,'' an OpenGL resource that allows one to render to off-screen textures. These framebuffers then need to have the required textures to render onto allocated as well. As each input and output from a pass can specify constraints on the texture's features, these constraints must be matched up for any connecting edges as well. Finally, in order to minimise memory usage we would like to re-use textures where possible. \\

Thus the allocation proceeds in three phases: reconciling texture constraints on edges between passes, computing how textures are shared between passes, and finally constructing all the necessary resources with the previously gathered information.

\subsection{Constraint Merging}\label{texture-constraints}
OpenGL textures include a massive amount of information.\cite{texparameter}\cite{teximage2d} When two ports are connected that specify different constraints on the texture properties, a join must be performed. Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error. For simplicity and brevity, we will focus on the join operator for a single texture property here, the internal format. The internal format is arguably the most important property. This property specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied. \\

The list of specified texture formats is quite large.\cite{texinternalformat} Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily. Instead, each format is represented by a constant, whose value has no relation to the features that format includes. This means that we need to first destructure each format name into a list of features: \\

@@

In other words, the algorithm proceeds backwards from the last node in the graph, marking predecessors' output ports with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at the nodes own output ports as available again. We repeat this process for each unique texture specification, each time ensuring we only consider ports that share that texture specification. \\

@@

Unfortunately print documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document. However, a brief animation of the pipeline illustrated in \autoref{fig:render-pipeline} is available online: \\

@@

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-shadow.png}
  \end{center}
  \caption{The depth map produced by the shadow map pass. It shows the distance of surfaces from the point of view of the primary light source.}
  \label{fig:pipeline-shadow}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.3\textwidth]{pipeline-blur.png}
  \end{center}
  \caption{The blurred high-pass from \autoref{fig:pipeline-lighting} produced by the gaussian blur pass. This is used to produce an effect called bloom in the final composite pass.}
  \label{fig:pipeline-blur}
\end{figure}

\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{pipeline-frame.png}
  \end{center}
  \caption{The final output frame that is produced by adding together the bloom texture from \autoref{fig:pipeline-blur} and the standard colour texture from \autoref{fig:pipeline-lighting}. Gamma adjustment and High Dynamic Range reduction are used for colour normalisation.}
  \label{fig:pipeline-frame}
\end{figure}

\section{Conclusion}\label{conclusion}
We have presented a system to handle the separation between objects that should be rendered, and the way in which they should be rendered. By using a protocol for the interaction between such objects and passes we are capable of controlling and combining a variety of rendering behaviour. With an additional protocol and set of algorithms we also present a high-level view of rendering pipelines, allowing users to conveniently compose effects to produce complicated scenes. \\

@@

Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes. This means that for instance an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it. Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass. This superclass then loosely defines the interaction. However, this does not solve issues of compatibility of shader code used by the object and the shader pass. In order to resolve this, we currently see two required features: \\

Shader source merging must be aware of uses-relations. This means that the code walker must track where variables, functions, and types are used and reorder the definitions and declarations in such a way that the dependencies are fulfilled. In GLSL, variables, functions, and types cannot be referenced before they are declared, requiring this reordering step. With this solved, shader code could be written in such a way to be more amiable towards combination and manipulation by external code. \\

GLSL's limited expressiveness makes it very difficult to figure out relations between code and to figure out how to properly combine pieces. A more high-level language as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components} should be introduced to allow a more convenient way to declare and write shaders, objects, and passes. With a higher-level language a compiler could more easily figure out how to combine features and reconcile differences between pieces of code. \\

@@

\section{Acknowledgements}\label{acknowledgements}
I would like to thank Michal ``phoe'' Herda, and all the people that will review this paper. Your name could be here!

>>>>

\begin{abstract}
  Modern real-time graphics make use of a lot of tricks in order to produce stunning visuals. Many of these tricks require separate rendering passes, as well as separate rendering logic for each pass. These passes are then combined in a variety of ways in order to produce a final image. The interaction between such a rendering pass and the objects it draws, as well as the interaction between multiple passes within a pipeline can become quite complex. Often times, in order to handle this complexity, the passes and objects are generalised, and the render logic is left almost entirely with either the object or the pass. We present a new method of representing objects, passes, and pipelines, which allows a modular encapsulation of effects and rendering behaviour, as well as object-oriented composition through inheritance. We make use of CLOS' multiple inheritance, multimethods, and standard method combination to form an extensible protocol. We also make use of the MOP, in order to introduce additional metadata to classes.
\end{abstract}

@@

\def\abovecaptionskip{1pt}
\def\listingautorefname{Listing}
\def\figureautorefname{Figure}

@@

Each of the customisable stages accepts only a single shader for each pass, making it difficult to separate, encapsulate, and ultimately combine behaviour. Furthermore, the steps required in order to change the shaders and shader inputs can be non-trivial and expensive to execute. \\

@@

This complexity results in a difficult challenge for modularity. We attempt to solve this challenge through several systems:

@@

The case study of the Unity game engine by Messaoudi et al\cite{unity} shows the availability of a set of fixed rendering pipelines that can be customised to a very limited extent with custom shaders. These shaders must fit into Unity's existing lighting and overall rendering model. While Unity does allow building a custom pipeline via their Scriptable Rendering Pipeline\cite{unitycustom}, they do not seem to offer any specific encapsulation or modularity features. \\

The work by He et al.\cite{components} introduces a framework for general encapsulation of shaders and their parameters into structures that minimise the overhead of changing GPU state while retaining the ability to dynamically compose shader parts. They do however not create a distinction between properties for rendering an object, and those for rendering an overall pass. \\

@@

In our previous work\cite{glsloop} we introduce a system to tie shader code to classes and to compose behaviour through inheritance. We make use of this system and extend it to allow further control over rendering behaviour in individual passes.

\section{Overview}\label{overview}
The system is composed of three distinct entities: shader objects, shader passes, and pipelines. Both shader objects and shader passes encapsulate rendering logic. The pipeline, on the other hand, represents the assembly of a complete frame, and only influences rendering in the sense that it tracks which passes to run in what order. \\

Due to the restrictions of the rendering pipeline of OpenGL, the combination of the render logic of objects and of passes is not trivial. We present a protocol to solve this difficulty in \hyperref[passes]{section 4}. \\

In \hyperref[pipelines]{section 5} we present a protocol for representing input and output information of a pass and for connecting these inputs and outputs between passes together. \\

@@

\section{Passes}\label{passes}
A shader pass should encapsulate the logic for drawing objects in a certain way. In order to accomplish its task, the pass needs to be able to partially or fully control the shaders of objects. For instance, an object should still be able to have control over how its vertices are constructed and what materials and textures are applied to it. The shader pass should be able to make use of this information if it needs to, or discard it completely if unneeded. \\

In order to permit this amount of control, we make use of the shader composition capabilities that we presented in our previous work\cite{glsloop}. This technique allows us to combine pieces of GLSL code. On top of this shader combination sits a new protocol to control the interaction between shader passes and objects. \\

@@

For any object that should be rendered using a given pass, first \code{register-object-for-pass} must be called. This call allows the pass to prepare the shader program that will be used during rendering for this object. A shader program is an OpenGL resource that compiles the shaders for the stages of a pass together. Using \\ \code{shader-program-for-pass} this program can then later be retrieved. The shader program is important for objects to access, as it allows them to set values for ``uniform'' variables that are used in the shader code. \\

\code{register-object-for-pass} calls \code{make-pass-shader-program} to compute this program, and registers it internally in the shader pass so that it can be retrieved later. If the shader pass would like to retain complete control over how each object is drawn, it would forego this call and instead generate its own shader program. Doing so is especially useful for post-processing effects that don't render any objects at all, and instead simply operate on textures that are output by previous passes. \\

@@

\code{coerce-pass-shader} computes the effective shader source for a particular shader stage or type. Typically this computation simply involves the combination of the shader sources of both the shader pass and of the object, for the given shader type. We perform this combination using the same parsing and code walking strategy as we described in our previous work.\cite{glsloop} \\

@@

This function is fine for allowing the object control over the behaviour. However, the pass cannot exert the same amount of control, due to the generic function's argument precedence. For example, if an \code{:around} method specialised on the object exists, a pass would not have any way of preventing it from firing, as its own \code{:around} method would be executed afterwards. We thus introduce another function with inverted argument order that is called first. \\

@@

An example of such a pass is illustrated in \autoref{lst:deferred-render}. The ports are modelled as slots of the class that carry additional metadata, requiring a new metaclass. A slot with a port-type will not only hold a texture for its value, but also retain information about how it is connected to other passes. Using slots in this manner also allows us to inherit ports from other classes, and thus combine behaviour alongside the shader source code. \\

@@

When assembling a pipeline, we can then connect these ports together, in order to dictate how the various output textures generated by a pass are used as inputs for other passes. As an example, the pipeline from \autoref{fig:render-pipeline} is created in \autoref{lst:pipeline}. \\

@@

\begin{step}
\item Create a pipeline object and instances of the desired shader passes.
\item Connect the ports of the shader passes.
\item Prepare the pipeline to allocate the needed resources.
\item Create instances of the desired objects.
\item Register each object with each pass.
\item Call \code{render} on the pipeline with a collection of all objects to draw.
\item The output texture of the output port of the last shader pass in the pipeline will contain the finished frame.
\end{step}

@@

\section{Allocation}\label{allocation}
In order to prepare the shader pipeline, several resources need to be allocated. Each shader stage needs a ``framebuffer,'' an OpenGL resource that allows one to render to off-screen textures. These framebuffers then need to have the required textures to render onto allocated as well. As each input and output from a pass can specify constraints on the features of the texture, these constraints must be matched up for any connecting edges as well. Finally, in order to minimise memory usage, we would like to re-use textures where possible. \\

Thus, the allocation proceeds in three phases: reconciling texture constraints on edges between passes, computing how textures are shared between passes, and finally constructing all the necessary resources with the previously gathered information.

\subsection{Constraint Merging}\label{texture-constraints}
OpenGL textures include a massive amount of information\cite{texparameter}\cite{teximage2d}. When two ports are connected that specify different constraints on the texture properties, a join must be performed. Fortunately, a wide range of the texture property values are fundamentally incompatible, meaning that a lot of the logic can simply error. For simplicity and brevity, we will focus on the join operator for a single texture property here: the internal format. The internal format is arguably the most important property. This property specifies how many colour channels the texture has, how many bits of precision each channel has, which format each channel has, as well as whether the texture is compressed or has sRGB gamma normalisation applied. \\

The list of specified texture formats is quite large\cite{texinternalformat}. Unfortunately OpenGL does not give us an interface to handle these formats in a way that lets us pick the individual features easily. Instead, each format is represented by a constant whose value has no relation to the features that format includes. This representation means that we first need to destructure each format name into a list of features: \\

@@

In other words, the algorithm proceeds backwards from the last node in the graph, marking output ports of predecessors with unique colours, then marking unconnected ports with unique colours, and finally marking all colours at the node's own output ports as available again. We repeat this process for each unique texture specification, each time ensuring we only consider ports that share that texture specification. \\

@@

Unfortunately printed documents cannot yet display animations, so illustrating the algorithm in motion is not possible directly in the document. However, a brief animation of the pipeline illustrated in \autoref{fig:render-pipeline} is available online: \\

@@

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.2\textwidth]{pipeline-shadow.png}
  \end{center}
  \caption{The depth map produced by the shadow map pass. It shows the distance of surfaces from the point of view of the primary light source.}
  \label{fig:pipeline-shadow}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.2\textwidth]{pipeline-blur.png}
  \end{center}
  \caption{The blurred high-pass from \autoref{fig:pipeline-lighting} produced by the gaussian blur pass. This is used to produce an effect called bloom in the final composite pass.}
  \label{fig:pipeline-blur}
\end{figure}

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.4\textwidth]{pipeline-frame.png}
  \end{center}
  \caption{The final output frame that is produced by adding together the bloom texture from \autoref{fig:pipeline-blur} and the standard colour texture from \autoref{fig:pipeline-lighting}. Gamma adjustment and High Dynamic Range reduction are used for colour normalisation.}
  \label{fig:pipeline-frame}
\end{figure}

\section{Conclusion}\label{conclusion}
We have presented a system to handle the separation between objects that should be rendered, and the way in which they should be rendered. By using a protocol for the interaction between such objects and passes, we are capable of controlling and combining a variety of rendering behaviour. With an additional protocol and set of algorithms we also present a high-level view of rendering pipelines, allowing users to conveniently compose effects to produce complicated scenes. \\

@@

Most severe is that there is no standardised protocol to communicate capabilities and data between objects and passes. This restriction means that for instance an object that does not possess texturing has no way of communicating this to a shader pass that might need to render it. Currently this problem is partially solved by introducing a mandatory superclass that objects need to be a subtype of, if they want to be renderable under a given pass. This superclass then loosely defines the interaction. However, this additional class does not solve issues of compatibility of shader code used by the object and by the shader pass. In order to properly resolve this restriction, we currently see two required features: \\

Shader source merging must be aware of uses-relations, meaning that the code walker must track where variables, functions, and types are used and reorder the definitions and declarations in such a way that the dependencies are fulfilled. In GLSL, variables, functions, and types cannot be referenced before they are declared, requiring this reordering step. With this problem solved, shader code could be written in such a way to be more amiable towards combination and manipulation by external code. \\

GLSL's limited expressiveness makes it very difficult to figure out relations between code and to figure out how to properly combine pieces. A more high-level language, as is used in other frameworks like Spark\cite{spark} or Shader Components\cite{components}, should be introduced to allow a more convenient way to declare and write shaders, objects, and passes. With a higher-level language, a compiler could more easily figure out how to combine features and reconcile differences between pieces of code. \\

@@

\section{Acknowledgements}\label{acknowledgements}
I would like to thank Michal ``phoe'' Herda, Robert Strandh, and all the people that will review this paper. Your name could be here!
